---
title: "ENV 790.30 - Time Series Analysis for Energy Data | Spring 2022"
subtitle: "Assignment 3 - Due date 02/08/22"
author: "Colin Lee"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: inline
---
  
## Directions
  
You should open the .rmd file corresponding to this assignment on RStudio. The file is available on our class repository on Github. 

Once you have the project open the first thing you will do is change "Student Name" on line 3 with your name. 
Then you will start working through the assignment by  **creating code and output** that answer each question. Be sure to use this assignment document. Your report should contain the answer to each question and any plots/tables you obtained (when applicable).

Please keep this R code chunk options for the report. It is easier for us to grade when we can see code and output together. And the tidy.opts will make sure that line breaks on your code chunks are automatically added for better visualization. 

```{r setup, include=FALSE} 
knitr::opts_chunk$set(echo = TRUE,tidy.opts=list(width.cutoff=80), tidy=FALSE) 
```

When you have completed the assignment, **Knit** the text and code into a single PDF file. Rename the pdf file such that it includes your first and last name (e.g., "LuanaLima_TSA_A03_Sp22.Rmd"). Submit this pdf using Sakai. 

## Questions 

Consider the same data you used for A2 from the spreadsheet "Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx". The data comes from the US Energy Information and Administration and corresponds to the January 2022 **Monthly** Energy Review. Once again you will work only with the following columns: Total Biomass Energy Production, Total Renewable Energy Production, Hydroelectric Power Consumption. Create a data frame structure with these three time series only.


R packages needed for this assignment:"forecast","tseries", and "Kendall". Install these packages, if you haven't done yet. Do not forget to load them before running your script, since they are NOT default packages.\\

```{r}
#Load/install required package here


library(forecast)
library(Kendall)
library(tseries)

library(lubridate)
library(ggplot2)
library(xlsx)
```

```{r dataframe}
#Importing data set

mydata <- read.xlsx(file = "/Users/colinlee/Documents/Duke/Spring 2022/ENV790/ENV790_TimeSeriesAnalysis_Sp2022/Data/Table_10.1_Renewable_Energy_Production_and_Consumption_by_Source.xlsx", sheetName = "Monthly Data", startRow = 12)

#cleaning data
mydata <- mydata[,4:6]

colnames(mydata)=c("Total Biomass Energy Production","Total Renewable Energy Production", "Hydroelectric Power Consumption")

head(mydata)
```

##Trend Component

### Q1
Create a plot window that has one row and three columns. And then for each object on your data frame, fill the plot window with time series plot, ACF and PACF. You may use the some code form A2, but I want all three plots on the same window this time. (Hint: use par() function)

```{r Q1}
par(mfrow=c(1,3))
plot(mydata[,"Total Biomass Energy Production"],type="l",col="red",ylab="Trillion Btu",xlab = "Month")
title(main="Time Series for Biomass Energy Production")
legend("topright",legend=c("Biomass Energy Production", "Mean Biomass Energy Production"), lty=c("solid","solid"),col=c("red","maroon"))

plot(mydata[,"Total Renewable Energy Production"],type="l",col="blue",ylab="Trillion Btu",xlab = "Month")
title(main="Time Series for Total Renewable Energy Production")
legend("topright",legend=c("Renewable Energy Production", "Mean Renewable Energy Production"), lty=c("solid","solid"),col=c("blue","violet"))

plot(mydata[,"Hydroelectric Power Consumption"],type="l",col="grey",ylab="Trillion Btu",xlab = "Month")
title(main="Time Series for Hydroelectric Power Consumption")
legend("topright",legend=c("Hydroelectric Power Consumption", "Mean Hydroelectric Power Consumption"), lty=c("solid","solid"),col=c("grey","black"))

par(mfrow=c(1,3))  #place plot side by side
acf(mydata[,1],lag.max=40,main=paste("Total Biomass Energy Production ACF"))
acf(mydata[,2],lag.max=40,main=paste("Total Renewable Energy Production ACF"))
acf(mydata[,3],lag.max=40,main=paste("Hydroelectric Power Consumption ACF"))

par(mfrow=c(1,3))  #place plot side by side
pacf(mydata[,1],lag.max=40,main=paste("Total Biomass Energy Production PACF"))
pacf(mydata[,2],lag.max=40,main=paste("Total Renewable Energy Production PACF"))
pacf(mydata[,3],lag.max=40,main=paste("Hydroelectric Power Consumption PACF"))
 
```


### Q2
From the plot in Q1, do the series Total Biomass Energy Production, Total Renewable Energy Production, Hydroelectric Power Consumption appear to have a trend? If yes, what kind of trend?

From the time series plots, it appears that Total Biomass Energy Production and Total Renewable Energy Production are generally increasing. However, it appears that Hydroelectric Power Consumption is generally decreasing.

Based on the ACF, it appears that Hydroelectric Power Consumption has a seasonal trend. Based on the ACFs for Total Biomass Energy Production and Total Renewable Energy Production, there is no seasonal trend, and the high autocorrelation indicates they are non-stationary. 

Additionally, we will need more tests to assess whether the time series are stochastic or deterministic.

We already know there is seasonality in Hydroelectric Power Consumption due to the clearer ACF, but the PACFs may indicate that there may be a little seasonality in Total Biomass Energy Production and Total Renewable Energy Production.

### Q3
Use the *lm()* function to fit a linear trend to the three time series. Ask R to print the summary of the regression. Interpret the regression output, i.e., slope and intercept. Save the regression coefficients for further analysis.

```{r Q3}
t <- c(1:nrow(mydata))
linear_trend_model_1=lm(mydata[,1]~t)
summary(linear_trend_model_1)
beta01=as.numeric(linear_trend_model_1$coefficients[1])  #first coefficient is the intercept term or beta0
beta11=as.numeric(linear_trend_model_1$coefficients[2])  #second coefficient is the slope or beta1

ggplot(mydata, aes(x = t, y=mydata[,1])) +
            geom_line(color="red") +
            ylab("Trillion Btu") +
            xlab("Month") +
            ggtitle("Total Biomass Energy Production") +
            geom_abline(intercept = beta01, slope = beta11, color="black")

linear_trend_model_2=lm(mydata[,2]~t)  
summary(linear_trend_model_2)
beta02=as.numeric(linear_trend_model_2$coefficients[1])  #first coefficient is the intercept term or beta0
beta12=as.numeric(linear_trend_model_2$coefficients[2])  #second coefficient is the slope or beta1

ggplot(mydata, aes(x = t, y=mydata[,2])) +
            geom_line(color="green") +
            ylab("Trillion Btu") +
            xlab("Month") +
            ggtitle("Total Renewable Energy Production") +
            geom_abline(intercept = beta02, slope = beta12, color="black")

linear_trend_model_3=lm(mydata[,3]~t) 
summary(linear_trend_model_3)
beta03=as.numeric(linear_trend_model_3$coefficients[1])  #first coefficient is the intercept term or beta0
beta13=as.numeric(linear_trend_model_3$coefficients[2])  #second coefficient is the slope or beta1

ggplot(mydata, aes(x = t, y=mydata[,3])) +
            geom_line(color="blue") +
            ylab("Trillion Btu") +
            xlab("Month") +
            ggtitle("Hydroelectric Power Consumption") +
            geom_abline(intercept = beta03, slope = beta13, color="black")
```

 
### Q4
Use the regression coefficients from Q3 to detrend the series. Plot the detrended series and compare with the plots from Q1. What happened? Did anything change?

Yes, we can see that the plots no longer have general trends of increasing or decreasing over time, but rather, remain flatter over long periods of time (save for the random/seasonal fluctuation). Additionally, this is shown through the linear models, where we can see the linear regression lines as completely flat where they were not in Q3.

In comparison to the plots from Q1, these plots no longer resemble the time series plots as well as these are more flatter in trajectory and no longer have increasing trends (biomass/renewable) or decreasing trend (hydro). Rather, they are more horizontal/flat and have no trend.

```{r Q4}
detrend_mydata_biomass <- mydata[,(1)]-(beta01+beta11*t)
detrend_mydata_renewable <- mydata[,(2)]-(beta02+beta12*t)
detrend_mydata_hydro <- mydata[,(3)]-(beta03+beta13*t)

ggplot(mydata, aes(x=t, y=detrend_mydata_biomass)) +
            geom_line(color="red") +
            ylab("Trillion Btu") +
            xlab("Month") +
            ggtitle("Total Biomass Energy Production") +
            geom_smooth(aes(y=detrend_mydata_biomass),color="black",method="lm") 

ggplot(mydata, aes(x=t, y=detrend_mydata_renewable)) +
            geom_line(color="green") +
            ylab("Trillion Btu") +
            xlab("Month") +
            ggtitle("Total Renewable Energy Production") +
            geom_smooth(aes(y=detrend_mydata_renewable),color="black",method="lm") 

ggplot(mydata, aes(x=t, y=detrend_mydata_hydro)) +
            geom_line(color="green") +
            ylab("Trillion Btu") +
            xlab("Month") +
            ggtitle("Hydroelectric Power Consumption") +
            geom_smooth(aes(y=detrend_mydata_hydro),color="black",method="lm") 
```


### Q5
Plot ACF and PACF for the detrended series and compare with the plots from Q1. Did the plots change? How?

The ACFs look very similar, with high seasonality for Hydroelectric Power Consumption and high autocorrelation among Total Biomass Energy Production and Total Renewable Energy Production. However, in the ACFs for Q5, it is evident that there might be some seasonality as there are some visible peaks and dips, especially in Total Renewable Energy Production. 

With regard to the PACFs, they remain very similar in appearance.

```{r Q5}
par(mfrow=c(1,3))  #place plot side by side
acf(detrend_mydata_biomass,lag.max=40,main=paste("Detrended Total Biomass Energy Production ACF"))
acf(detrend_mydata_renewable,lag.max=40,main=paste("Detrended Total Renewable Energy Production ACF"))
acf(detrend_mydata_hydro,lag.max=40,main=paste("Detrended Hydroelectric Power Consumption ACF"))

par(mfrow=c(1,3))  #place plot side by side
pacf(detrend_mydata_biomass,lag.max=40,main=paste("Detrended Total Biomass Energy Production PACF"))
pacf(detrend_mydata_renewable,lag.max=40,main=paste("Detrended Total Renewable Energy Production PACF"))
pacf(detrend_mydata_hydro,lag.max=40,main=paste("Detrended Hydroelectric Power Consumption PACF"))
```


## Seasonal Component

Set aside the detrended series and consider the original series again from Q1 to answer Q6 to Q8.

### Q6
Do the series seem to have a seasonal trend? Which serie/series? Use function *lm()* to fit a seasonal means model (i.e. using the seasonal dummies) to this/these time series. Ask R to print the summary of the regression. Interpret the regression output. Save the regression coefficients for further analysis.

I really only think hydro has seasonality, but I will do this for all the series just to see in case I'm wrong!

```{r Q6}
ts_mydata <- ts(mydata, start = 1, frequency = 12)

# hydro
dummies_hydro <- seasonaldummy(ts_mydata[,3])
seas_means_model_hydro=lm(ts_mydata[,3]~dummies_hydro)
summary(seas_means_model_hydro)
beta_int_hydro=seas_means_model_hydro$coefficients[1]
beta_coeff_hydro=seas_means_model_hydro$coefficients[2:12]

# biomass
dummies_biomass <- seasonaldummy(ts_mydata[,1])
seas_means_model_biomass=lm(ts_mydata[,1]~dummies_biomass)
summary(seas_means_model_biomass)
beta_int_biomass=seas_means_model_biomass$coefficients[1]
beta_coeff_biomass=seas_means_model_biomass$coefficients[2:12]

# renewable
dummies_renew <- seasonaldummy(ts_mydata[,2])
seas_means_model_renew=lm(ts_mydata[,2]~dummies_renew)
summary(seas_means_model_renew)
beta_int_renew=seas_means_model_renew$coefficients[1]
beta_coeff_renew=seas_means_model_renew$coefficients[2:12]
```

Based on the summary data, it is clear that only the Hydroelectric Power Consumption data is seasonal because of a low p-value of 2.2E-16. The Total Biomass Energy Production is most definitely not seasonal due to a very high p-value of 0.8647, and this was somewhat apparent with the detrended ACF. For the Total Renewable Energy Production, the p-value is relatively low at 0.07, indicating that there may be some seasonality, but this is not statistically significant for conventional p-value of 0.05. 

### Q7
Use the regression coefficients from Q6 to deseason the series. Plot the deseason series and compare with the plots from part Q1. Did anything change?

I will not proceed with Total Biomass Energy Production or Total Renewable Energy Production due to insignificant levels of seasonality.

Compared to the time series plot in Q1 for Hydroelectric Power Consumption, the scale is much lower with ranges oscillating over zero rather than between 200-300. Furthermore, there are less serious oscillations as there are less big peaks and dips that are associated with the annual seasonality, making the this deseasoned version less wavy as the Q1 time series plot.

```{r Q7}
hydro_seas_comp=array(0,nrow(mydata))
for(i in 1:nrow(mydata)){
  hydro_seas_comp[i]=(beta_int_hydro+beta_coeff_hydro%*%dummies_hydro[i,])
}

deseason_hydro <- mydata[,3]-hydro_seas_comp

ggplot(mydata, aes(x=t, y=deseason_hydro)) +
            geom_line(color="blue") +
            ylab("Hydroelectric Power Consumption") +
            xlab("Month") +
            ggtitle("Deseasoned Hydroelectric Power Consumption")
```


### Q8
Plot ACF and PACF for the deseason series and compare with the plots from Q1. Did the plots change? How?

The ACF changed significantly, where the ACF for hydro in Q1 is very seasonal and very hard to assess for autocorrelation. Now, deseasoned, the ACF no longer has this wave pattern, but more resembles a decreasing linear slope.

The PACF is slightly different in that the y-axis range for PACF values is much lower for the deseasoned hydro data compared to the Q1 hydro data (no adjustments made). With regards to changes in how the PACFs look, it is hard to tell any significant differences besides the change in range.

```{r Q8}
acf(deseason_hydro,lag.max=40,main=paste("Deseasoned Hydroelectric Power Consumption ACF"))
pacf(deseason_hydro,lag.max=40,main=paste("Deseasoned Hydroelectric Power Consumption PACF"))
```



